Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-07 results.tex
pass after rearrangement
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@955 ccdd05d5-56c1-461a-bf98-85f6876523b4

We examine the performance of the \lloid\ algorithm on a small region of
compact binary parameter space centered on typical NS-NS masses.  We begin by
constructing a template bank that spans component masses between
1~and~3~$M_\odot$ using a simulated advanced LIGO noise curve~\cite{lal}.  Then
we create sub banks based on chirp mass slices in that parameter space.
Figure{fig:tmpltbank} illustrates this procedure.
\begin{figure}[htbp]
	\label{fig:tmpltbank}
	\begin{center}
		\includegraphics{figures/tmpltbank.png}
		\caption{Placement of template parameters used in this paper.  The template bank consists of 98544 templates with component masses $m_1$, $m_2$, between 1~and~3~$M_\odot$.  We shall design a filter bank to search for a small subset of these template parameters with chirp masses $\mathcal M$ between 1.1955~and~1.2045~$M_\odot$.}
	\end{center}
\end{figure}
The result is that we obtain 657 templates with chirpmasses between 1.1955~and~1.2045~$M_\odot$.  With this
sub-bank we were able to construct an efficient time slice decomposition that consisted of 11 time slices
with sample rates between 32 and 4096 Hz shown in table \ref{tab:time_slices}.
\begin{table}
\begin{indented}
\caption{\label{tab:time_slices} Filter design for these 657 templates.  From left to right, this table shows the sample rate, time interval, number of samples, and number of orthogonal templates for each time slice.  The \SVD{} tolerance is varied from $\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.}
\item[]\begin{tabular}{rr@{,}lc*{6}{r}}
\br
\multicolumn{4}{c}{} &\centre{6}{$\log_{10}$ (1 - \textsc{svd} tolerance)} \\
\ns
\multicolumn{4}{c}{} &\crule{6} \\
$f^s$ (Hz) & $[t^s$&$t^{s+1})$ (s) & $N^s$ & -1 & -2 & -3 & -4 & -5 & -6 \\
\mr
4096 & [0&0.5) & 2048 & 1 & 4 & 6 & 8 & 10 & 14 \\
512 & [0.5&4.5) & 2048 & 2 & 6 & 8 & 10 & 12 & 16 \\
256 & [4.5&12.5) & 2048 & 2 & 6 & 8 & 10 & 12 & 15 \\
128 & [12.5&76.5) & 8192 & 6 & 20 & 25 & 28 & 30 & 32 \\
64 & [76.5&140.5) & 4096 & 1 & 8 & 15 & 18 & 20 & 22 \\
64 & [140.5&268.5) & 8192 & 1 & 7 & 21 & 25 & 28 & 30 \\
64 & [268.5&396.5) & 8192 & 1 & 1 & 15 & 20 & 23 & 25 \\
32 & [396.5&460.5) & 2048 & 1 & 1 & 3 & 9 & 12 & 14 \\
32 & [460.5&588.5) & 4096 & 1 & 1 & 7 & 16 & 18 & 21 \\
32 & [588.5&844.5) & 8192 & 1 & 1 & 8 & 26 & 30 & 33 \\
32 & [844.5&1100.5) & 8192 & 1 & 1 & 1 & 12 & 20 & 23 \\
\br
\end{tabular}
\end{indented}
\end{table}
We use this template bank and decomposition for the remainder of this section.

@@

We expect two known contributions to the \SNR\ loss to arise in our
implementation of the \lloid\ algorithm.  The first is the \SNR\ loss due to
the truncation of the \SVD\ basis and estimates for it
exist~\cite{Cannon:2010p10398}.  The second comes from non ideal
implementations of resampling that cause signal loss.  The \SNR\ loss is to be compared
with the mismatch of 0.03 that arises from the discreteness of the
template bank which is typically.  We will consider an acceptable
\SNR\ loss to be a factor of 10 smaller than this, i.e., no more than $\sim$0.003.

@@

In this section we check the effect of the \SVD\ tolerance parameter defined
in~\cite{Cannon:2010p10398}.  By changing the tolerance of the reconstruction of
the physical waveforms one effects directly the mismatch that results from the
truncation of the orthogonal filter matrix.  The conditions presented here are
more complicated than in the original work~\cite{Cannon:2010p10398} due to the
inclusion of the time sliced templates and resampling.  \FIXME{However, we find
that the results are quite as expected.}  Figure \ref{fig:hist-svd-tolerance}
demonstrates that it is possible to acheive typical \SNR\ losses of $\ll1\%$.
However, we do notice that the mismatch saturates with respect to the tolerance.
This could be the effect of the resampling, or another \SNR\ loss that we did
not model or expect.  However the saturation is still an order of magnitude below our target mismatch of $\sim$0.003.  We find that an \SVD\ tolerance of
$\left(1-10^{-4}\right)$ is adequate. 
%
\begin{figure}
	\begin{center}
		\subfloat[Mismatch versus \textsc{svd} tolerance]{\label{fig:hist-svd-tolerance}\includegraphics{bw.pdf}}
		\subfloat[Mismatch versus $N^\shortuparrow$]{\label{fig:hist-interpolate}\includegraphics{bw_resample.pdf}}
		\caption{Box-and-whisker plot of mismatch between nominal
template bank and \lloid\ measured impulse respones.  The upper and lower boundaries of
the boxes show the upper and lower quartiles; the lines in the center denote the medians.
The whiskers represent the minimum and maximum mismatch over all templates.  In 
\subref{fig:hist-svd-tolerance} the interpolation filter length is held fixed
at $N^\shortuparrow = 192$, while the \SVD\ tolerance is varied from
$\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.  In \subref{fig:hist-interpolate}, the \SVD\ tolerance is fixed at $\left(1-10^{-6}\right)$ while $N^\shortuparrow$ is varied from 8 to 192 coefficients.}
	\end{center}
\end{figure}

@@

\Table{\label{table:flops}Computational cost in \flops\ of the \TD\ method, the \FD\ method, and \lloid\ for the bank of $2 \times 657$ templates sampled at 4096 Hz for chirp masses $1.1955 \leqslant \mathcal{M} / M_\odot \leqslant 1.2045$.  A block size of $D = 2 N$ is used for the \fft{}s in the \FD\ method.  Both the \FD\ method and \lloid\ are five orders of magnitude faster than the conventional \TD\ method.  However, the \FD\ method has a latency of over half of an hour, whereas \lloid\ has no latency at all.}
\br
Method & \flops\ & latency (s) \\
\mr
Time-domain (\TD) convolution & $2.4\times10^{13}$ & 0 \\
Frequency-domain (\FD) convolution & $2.6\times10^8$ & 2201 \\
\lloid\ (this work) & $4.7\times10^8$ & 0 \\
\br
\end{tabular}
\end{indented}
\end{table}

\subsubsection{Extrapolation of the computational cost to an Advanced \LIGO{} search}
\editorial{Drew: I think this section be the last section of the results section. Should we move it to be Sec. 4.2? Nick: I'm inclined to agree with Drew, though I see that we want to make explicit that it's an analysis of the algorithm and not of the implementation and the current placement aids that. Also, the given FLOPS numbers correspond to what choice of \SVD{} tolerance and $N^\shortuparrow$?}

Table~\ref{table:flops} shows that the \lloid\ method requires $\sim 10^8$ \flops\
to cover a sub-bank of comprising $\sim 10^2$ out of $\sim 10^5$ sets of intrinsic
source parameters.  Given that modern (ca. 2011) workstations can sustain
computation rates up to $\sim 10^{10}$ \flops{}, and assuming that other regions of the
parameter space have similar computational scaling, an entire search could be implemented
with $\gtrsim 10$ machines.

>>>>

We examine the performance of the \lloid\ algorithm on a small region of
compact binary parameter space centered on typical NS-NS masses.  We begin by
constructing a template bank that spans component masses between
1~and~3~$M_\odot$ using a simulated advanced LIGO noise curve~\cite{lal}.  Then
we create sub banks based on chirp mass slices in that parameter space.
Figure \ref{fig:tmpltbank} illustrates this procedure.
\begin{figure}[htbp]
	\begin{center}
		\includegraphics{figures/tmpltbank.png}
		\caption{\label{fig:tmpltbank}Placement of template parameters used in this paper.  The template bank consists of 98544 templates with component masses $m_1$, $m_2$, between 1~and~3~$M_\odot$.  We shall design a filter bank to search for a small subset of these template parameters with chirp masses $\mathcal M$ between 1.1955~and~1.2045~$M_\odot$.}
	\end{center}
\end{figure}
The result is that we obtain 657 templates with chirpmasses between 1.1955~and~1.2045~$M_\odot$.  With this
sub-bank we were able to construct an efficient time slice decomposition that consisted of 11 time slices
with sample rates between 32 and 4096 Hz shown in table \ref{tab:time_slices}.
\begin{table}
\begin{indented}
\caption{\label{tab:time_slices} Filter design for these 657 templates.  From left to right, this table shows the sample rate, time interval, number of samples, and number of orthogonal templates for each time slice.  The \SVD{} tolerance is varied from $\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.}
\item[]\begin{tabular}{rr@{,}lc*{6}{r}}
\br
\multicolumn{4}{c}{} &\centre{6}{$\log_{10}$ (1 - \textsc{svd} tolerance)} \\
\ns
\multicolumn{4}{c}{} &\crule{6} \\
$f^s$ (Hz) & $[t^s$&$t^{s+1})$ (s) & $N^s$ & -1 & -2 & -3 & -4 & -5 & -6 \\
\mr
4096 & [0&0.5) & 2048 & 1 & 4 & 6 & 8 & 10 & 14 \\
512 & [0.5&4.5) & 2048 & 2 & 6 & 8 & 10 & 12 & 16 \\
256 & [4.5&12.5) & 2048 & 2 & 6 & 8 & 10 & 12 & 15 \\
128 & [12.5&76.5) & 8192 & 6 & 20 & 25 & 28 & 30 & 32 \\
64 & [76.5&140.5) & 4096 & 1 & 8 & 15 & 18 & 20 & 22 \\
64 & [140.5&268.5) & 8192 & 1 & 7 & 21 & 25 & 28 & 30 \\
64 & [268.5&396.5) & 8192 & 1 & 1 & 15 & 20 & 23 & 25 \\
32 & [396.5&460.5) & 2048 & 1 & 1 & 3 & 9 & 12 & 14 \\
32 & [460.5&588.5) & 4096 & 1 & 1 & 7 & 16 & 18 & 21 \\
32 & [588.5&844.5) & 8192 & 1 & 1 & 8 & 26 & 30 & 33 \\
32 & [844.5&1100.5) & 8192 & 1 & 1 & 1 & 12 & 20 & 23 \\
\br
\end{tabular}
\end{indented}
\end{table}
We use this template bank and decomposition for the remainder of this section.

@@

We expect two known contributions to the \SNR\ loss to arise in our
implementation of the \lloid\ algorithm.  The first is the \SNR\ loss due to
the truncation of the \SVD\ basis and estimates for it
exist~\cite{Cannon:2010p10398}.  The second comes from non ideal
implementations of resampling that cause signal loss.  The \SNR\ loss is to be
compared with the mismatch of 0.03 that arises from the discreteness of the
template bank.  We will consider a target \SNR\ loss to be a factor of 10
smaller than this, i.e., no more than $\sim$0.003.

@@

In this section we check the effect of the \SVD\ tolerance parameter defined
in~\cite{Cannon:2010p10398}.  By changing the tolerance of the reconstruction
of the physical waveforms one effects directly the mismatch that results from
the truncation of the orthogonal filter matrix.  The conditions presented here
are more complicated than in the original work~\cite{Cannon:2010p10398} due to
the inclusion of the time sliced templates and resampling.  Figure
\ref{fig:hist-svd-tolerance} demonstrates that it is possible to acheive
typical \SNR\ losses of $\ll1\%$.  However, we do notice that the mismatch
saturates with respect to the tolerance.  This could be the effect of the
resampling, or another \SNR\ loss that we did not model or expect.  However the
saturation is still an order of magnitude below our target mismatch of
$\sim$0.003.  We find that an \SVD\ tolerance of $\left(1-10^{-4}\right)$ is
adequate to acheive our target SNR loss.
%
\begin{figure}
	\begin{center}
		\subfloat[Mismatch versus \textsc{svd} tolerance]{\label{fig:hist-svd-tolerance}\includegraphics{bw.pdf}}
		\subfloat[Mismatch versus $N^\shortuparrow$]{\label{fig:hist-interpolate}\includegraphics{bw_resample.pdf}}
		\caption{Box-and-whisker plot of mismatch between nominal
template bank and \lloid\ measured impulse respones.  The upper and lower boundaries of
the boxes show the upper and lower quartiles; the lines in the center denote the medians.
The whiskers represent the minimum and maximum mismatch over all templates.  In 
\subref{fig:hist-svd-tolerance} the interpolation filter length is held fixed
at $N^\shortuparrow = 192$, while the \SVD\ tolerance is varied from
$\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.  In \subref{fig:hist-interpolate}, the \SVD\ tolerance is fixed at $\left(1-10^{-6}\right)$ while $N^\shortuparrow$ is varied from 8 to 192 coefficients.}
	\end{center}
\end{figure}

@@

We are now prepared to offer the estimated computational cost of filtering this
sub-bank of templates compared to other methods.  We used the results of the
previous sub sections to set the \SVD\ tolerance to $1-10^{-4}$  and the
interpolation filter length to 16.  We find that the \lloid\ method provides
comparable performance to the standard \FD\ approach but with no intrinsic
latency.  Table \ref{table:flops} lists the costs for filtering this sub-bank.
%
\Table{\label{table:flops}Computational cost in \flops\ of the \TD\ method, the \FD\ method, and \lloid\ for the bank of $2 \times 657$ templates sampled at 4096 Hz for chirp masses $1.1955 \leqslant \mathcal{M} / M_\odot \leqslant 1.2045$.  A block size of $D = 2 N$ is used for the \fft{}s in the \FD\ method.  Both the \FD\ method and \lloid\ are five orders of magnitude faster than the conventional \TD\ method.  However, the \FD\ method has a latency of over half of an hour, whereas \lloid\ has no latency at all.}
\br
Method & \flops\ & latency (s) \\
\mr
Time-domain (\TD) convolution & $2.4\times10^{13}$ & 0 \\
Frequency-domain (\FD) convolution & $2.6\times10^8$ & 2201 \\
\lloid\ (this work) & $4.7\times10^8$ & 0 \\
\br
\end{tabular}
\end{indented}
\end{table}

\subsection{Extrapolation of the computational cost to an Advanced \LIGO{} search}

Table~\ref{table:flops} shows that the \lloid\ method requires $\sim 10^8$
\flops\ to cover a sub-bank comprising of $\sim 10^2$ out of the total $\sim
10^5$ mass pairs.  Given that modern (ca. 2011) workstations can sustain
computation rates up to $\sim 10^{10}$ \flops{}, and assuming that other
regions of the parameter space have similar computational scaling\footnote{The
length of templates varies over the parameter space.  Lower mass templates are
longer and would require more computational cost, higher mass templates are
shorter and would require less.  We consider the estimates based on this
sub-bank to be a reasonable representation, however}, an entire search could be
implemented with $\gtrsim 10$ machines.
